Bandwidth-Aware Reconfigurable Memory Design with Hybrid Memory Technologies

### 2011-04-09 paper.tex

While a number techniques can be found in today's systems and research work to
alleviate such bandwidth bottleneck, few can improve the system throughput in a
power and cost efficient manner. High performance computing machines such as
NVIDIA's Tesla~\cite{nvidia:tesla} rely on very high main memory bandwidths
(provided by GDDR memories) to feed the requirement from the processor. However,
GDDR memories run at higher clock rates and are more power hungry than
conventional DRAM modules. It is undesirable for either general purpose or high
performance computing systems to improve their computing performance by
sacrifacing power efficiency. With the emerging 3D chip technology, caches can
be stacked on top of processor cores to provide high memory
bandwidth[12][21][22]. However, ...

Caching has long been employed as the most effective approach to reduce memory
access latency. Adding an extra level of cache, however, can also help alleviate
the increasing bandwidth pressure of off-chip memory. In [dac.1] the authors
discuss extensively the problem of limited pin bandwidth to multiprocessor
systems. They have focused on program performance in multiprocessor systems and
make detailed decomposition of program execution times. They conclude that even
more complex on-chip cache structures would prove to be costeffective, with the
limited pin bandwidth severely restricting performance increases. In [dac.2] the
authors have carefully studied the requirements for on-chip cache structures
along with all sorts of optimization techniques that come along with scaling of
processor core numbers. Their study shows that cache size need to grow much
faster than processor core numbers to compensate for the limited off-chip
bandwidth. Because of that, the near future processors need to allocate a huge
percentage of chip area for caches, which means much less core counts than
expected. The study also shows that effective bandwidth optimization techniques
can help reduce cache size requirement and thus help scaling processor cores.
However, the effectiveness of caching in the mid/large scale multiprocessor
systems could be highly suboptimal due to significant contention across the
parallel tasks. The more bandwidth requirement by a workload can be satisfied by
the last level cache~(LLC), the less bandwidth will be required from the
off-chip main memory. Our memory technology pre-exploration shows that the
bandwidth provided by a memory decreases with the increase of its capacity.
Based on these observations, adding more levels of cache can potentially fill in
the bandwidth gap.

On-chip caches with fast random access, high storage density, and non-volatility
become possible due to the emergence of various new non-volatile memory~(NVM)
technologies, such as spin-torque-transfer memory~(STT-RAM), phase-change
memory~(PCRAM), and resistive memory~(RRAM). These emerging memory technologies
are believed to be promising solution as on-chip
caches~\cite{Sun:2009:MRAM-L2-CMP}. In addition to the benefit of
non-volatility, we demonstrate that these memory technologies provide higher
bandwidth than SRAM at a high capacity. Consequently, it is beneficial to use
NVM as lower level caches when large capacity.

@@

The bandwidth problem of multicore processors has drawn much attention recently.
Yu \emph{et al.} proposed a LLC partitioning algorithm to minimize bandwidth
requirement to off-chip main memory~\cite{Yu:2010:OCMB}. While most cache
partitioning techniques focus on cache miss rates, our work takes a different
approach in which tasks¡¯ memory bandwidth requirements are taken into account
when identifying a cache partitioning for multi-programmed and/or multithreaded
workloads. Cache resources are allocated with the objective that the overall
system bandwidth requirement is minimized for the target workload. The key
insight is that cache miss-rate information may severely misrepresent the actual
bandwidth demand of the task, which ultimately determines the overall system
performance and power consumption. <the moguls paper, the figure 1, however design for
induvidual application, i.e., with different application can generate different
designs.>

@@

Based on our memory technology exploration, we now provide an overview of BAROM
design. While many design methods envolved with new memory technologies endeavor
to reduce the off-chip memory access latency, this work focuses on minimize
off-chip bandwidth requirement by employing hybrid on-chip memory and
reconfiguration. For each level of memory, we select the memory technology that
has the highest provided bandwidth within given capacity. The total capacity of
each level is then partitioned to subsets of ways with different provided
bandwidths. On top of such hardware hierarchy design, we apply dynamic memory
system reconfiguration with the bandwidth demand of a specific set of
applications as the major metric. One desirable characteristic of BAROM is that
it provides prioritized access to the residual memory capacity available at
runtime instead of sharing it among all memory accesses. By having control over
the residual capacity, we can better utilize the limited resources to enhance
system performance. The hardware reconfiguration is further facilitated by a
prediction engine. Rather than conventional last value or history table based
predictors, we employ a probablity-based statistical predictor which can achieve
higher accuracy with small performance overhead. The rest of this section
describes the BAROM design, which is composed of on-chip memory hierarchy
design, the reconfiguration mechanism, and the prediction engine.

@@

Memory reconfiguration relies on accurate predictions of demand bandwidth of a
workload in response to the dynamically varying application characteristics.
Conventional last value or table based predictors can model neither long range
patterns of an application nor patterns with variable lengths. In this work, we
employ a statistical predictor to support for the BAROM design. The basic idea
of the predictor is borrowed from the n-Gram models, which are typically used by
speech and natural language processing~\cite{smoothing-tech}. The n-Gram models
are usually formulated as a probability distribution $p(s)$ over a strings $s$,
and attemp to reflect how frequently the string occurs as a sentense. Prediction
on a reasonable combination of strings is then generated based on the
probability distribution. Our predictor employs the same basic idea, but
simplifies the model for performance considerations. In our model, each demand
bandwidth sample abtained in a time interval is a string, and a given length
(the ``order'') of samples is stored in a table as a pattern. We calculate the
probability of each pattern at each time interval, and predict the demand
bandwidth for the next time interval based on the probabilities.

>>>>

While a number techniques can be found in today's systems and research work to
alleviate such bandwidth bottleneck, few can improve the system throughput in a
power and cost efficient manner. High performance computing machines such as
NVIDIA's Tesla~\cite{nvidia:tesla} rely on very high main memory bandwidths
(provided by GDDR memories) to feed the requirement from the processor. However,
GDDR memories run at higher clock rates and are more power hungry than
conventional DRAM modules. It is undesirable for either general purpose or high
performance computing systems to improve their computing performance by
sacrificing power efficiency. With the emerging 3D chip technology, caches can
be stacked on top of processor cores to provide high memory
bandwidth[12][21][22]. However, ...

Caching has long been employed as the most effective approach to reduce memory
access latency. Proper on-chip memory hierarchy design, however, can also help
alleviate the increasing bandwidth pressure of off-chip memory. In [dac.1] the
authors discuss extensively the problem of limited pin bandwidth to
multiprocessor systems. They have focused on program performance in
multiprocessor systems and make detailed decomposition of program execution
times. They conclude that even more complex on-chip cache structures would prove
to be costeffective, with the limited pin bandwidth severely restricting
performance increases. In [dac.2] the authors have carefully studied the
requirements for on-chip cache structures along with all sorts of optimization
techniques that come along with scaling of processor core numbers. Their study
shows that cache size need to grow much faster than processor core numbers to
compensate for the limited off-chip bandwidth. Because of that, the near future
processors need to allocate a huge percentage of chip area for caches, which
means much less core counts than expected. The study also shows that effective
bandwidth optimization techniques can help reduce cache size requirement and
thus help scaling processor cores. However, the effectiveness of caching in the
mid/large scale multiprocessor systems could be highly suboptimal due to
significant contention across the parallel tasks. The more bandwidth requirement
by a workload can be satisfied by the last level cache~(LLC), the less bandwidth
will be required from the off-chip main memory. Our memory technology
pre-exploration shows that the bandwidth provided by a memory decreases with the
increase of its capacity. Based on these observations, adding more levels of
cache can potentially fill in the bandwidth gap.

On-chip caches with fast random access, high storage density, and non-volatility
become possible due to the emergence of various new non-volatile memory~(NVM)
technologies, such as spin-torque-transfer memory~(STT-RAM), phase-change
memory~(PCRAM), and resistive memory~(RRAM). These emerging memory technologies
are believed to be promising solution as on-chip
caches~\cite{Sun:2009:MRAM-L2-CMP}. In addition to the benefit of
non-volatility, we demonstrate that these memory technologies provide higher
bandwidth than SRAM at a large capacity. Consequently, it is beneficial to use
NVM as lower level caches when large capacity.

@@

The bandwidth problem of multicore processors has drawn much attention recently.
Yu \emph{et al.} proposed a LLC partitioning algorithm to minimize bandwidth
requirement to off-chip main memory~\cite{Yu:2010:OCMB}. While most cache
partitioning techniques focus on cache miss rates, their work takes a different
approach in which tasks¡¯ memory bandwidth requirements are taken into account
when identifying a cache partitioning for multi-programmed and/or multithreaded
workloads. Cache resources are allocated with the objective that the overall
system bandwidth requirement is minimized for the target workload. The key
insight is that cache miss-rate information may severely misrepresent the actual
bandwidth demand of the task, which ultimately determines the overall system
performance and power consumption. <the moguls paper, the figure 1, however design for
induvidual application, i.e., with different application can generate different
designs.>

@@

Based on our memory technology exploration, we now provide an overview of BAROM
design. While many design methods involved with new memory technologies endeavor
to reduce the off-chip memory access latency, this work focuses on minimizing
off-chip bandwidth requirement by employing hybrid on-chip memory and
reconfiguration. For each level of memory, we select the memory technology that
has the highest provided bandwidth within given capacity. The total capacity of
each level is then partitioned to subsets of ways with different provided
bandwidths. On top of such hardware hierarchy design, we apply dynamic memory
system reconfiguration with the bandwidth demand of a specific set of
applications as the major metric. One desirable characteristic of BAROM is that
it provides prioritized access to the residual memory capacity available at
runtime instead of sharing it among all memory accesses. By having control over
the residual capacity, we can better utilize the limited resources to enhance
system performance. The hardware reconfiguration is further facilitated by a
prediction engine. Rather than conventional last value or history table based
predictors, we employ a probability-based statistical predictor which can
achieve higher accuracy with small performance overhead. The rest of this
section describes the BAROM design, which is composed of on-chip memory
hierarchy design, the reconfiguration mechanism, and the prediction engine.

@@

Memory reconfiguration relies on accurate predictions of demand bandwidth of a
workload in response to the dynamically varying application characteristics.
Conventional last value or table based predictors can model neither long range
patterns of an application nor patterns with variable lengths. In this work, we
employ a statistical predictor to support for the BAROM design. The basic idea
of the predictor is borrowed from the n-Gram models, which are typically used by
speech and natural language processing~\cite{smoothing-tech}. The n-gram models
are usually formulated as a probability distribution $p(s)$ over a strings $s$,
and attemp to reflect how frequently the string occurs as a sentense. Prediction
on a reasonable combination of strings is then generated based on the
probability distribution. Our predictor employs the same basic idea, but
simplifies the model for performance considerations. In our model, each demand
bandwidth (DBW) sample abtained in a time interval is analogous to a string in
language, and a given length (the ``order'') of samples is stored in a table as
a pattern. At each time interval, the prediction engine will update the pattern
table with the new DBW sample, and calculate probability of the updated pattern.
The probability of a pattern $s$ of length $l$ can be calculated without loss as
the product of a sequence of conditional probabilities using the equation shown
below.

\begin{equation}
  p(s)=p(w_1)p(w_2|w_1)...p(w_l|w_1...w_{l-1})=\prod_{i=1}^{l}p(w_i|w_1...w_{i-1})\label{eq:pr1}
\end{equation}
where $w_l$ is the DBW sample obtained in current time interval, and $w_1$
through $w_{l-1}$ are the preceding $l-1$ DBW samples. In n-gram models, we make
the approximation that each conditional probability only depends on the
preceding $n-1$ samples and obtain the following equation.

\begin{equation}
  p(s)=\prod_{i=1}^{l}p(w_i|w_{i-n+1}^{i-1})\label{eq:pr2}
\end{equation}
in which $w_{i-n+1}^{i-1}$ denotes the sequence of $w_{i-n+1}...w_{i-1}$. In
order to compute the result of Equation~\ref{eq:pr2}, an estimation of of
$p(w_i|w_{i-n+1}^{i-1})$ can be generated using maximum likelihood~(ML)
estimation method. In n-gram models, the largest n in wide use is n=3 which is
called a trigram model. Therefore, we adopt n=3 in our predictor. Each
conditional probability is calculated using ML estimation in
Equation~\ref{eq:pr3}.

\begin{equation}
  \hat{p}(w_i|w_{i-n+1}^{i-1})=\frac{c(w_{i-2}^{i})}{c(w_{i-2}^{i-1})}\label{eq:pr3}
\end{equation}
where $c(w_{a}^{b})$ is the number of times that the sequence $w_a..w_b$ appears
in preceding samples.

The demand bandwidth in the next time interval is predicted based on the
probabilities.
